---
title: "Assignment 1 - February 10th, 2026"
subtitle: EPPS 6323 Knowledge Mining
format: html
---

# R Programming Basic Commands

## Lab 01

### Create object using the assignment operator (\<-, =)

```{r}
x <- c(1,3,2,5)
x
x = c(1,6,2)
x
y = c(1,4,3)
```

### Using function

```{r}
length(x)  # What does length() do?
length(y)
```

### Using +, -, \*, /,\^ operators

```{r}
x+y
ls() # List objects in the environment
rm(x,y) # Remove objects
ls()
rm(list=ls()) # Danger! What does this do?  Not recommended!
```

### Matrix operations

```{r}
?matrix
x=matrix(data=c(1,2,3,4), nrow=2, ncol=2) # Create a 2x2 matrix object
x
x=matrix(c(1,2,3,4),2,2)
matrix(c(1,2,3,4),2,2,byrow=T) # What about byrow=F?

sqrt(x) # What does x look like?
x
x^2

x=rnorm(50) # Generate a vector of 50 numbers using the rnorm() function

y=x+rnorm(50,mean=50,sd=.1) # What does rnorm(50,mean=50,sd=.1) generate?

cor(x,y) # Correlation of x and y
set.seed(1303) # Set the seed for Random Number Generator (RNG) to generate values that are reproducible.
rnorm(50)
set.seed(3) # Try different seeds?
y=rnorm(100)
```

### Simple descriptive statistics (base)

```{r}
mean(y)
var(y)
sqrt(var(y))
sd(y)
```

### Visualization using R Graphics (without packages)

```{r}
x=rnorm(100)
y=rnorm(100)
plot(x,y)

plot(x,y, pch=20, col = "firebrick") # Scatterplot for two numeric variables by default
plot(x,y, pch=20, col = "steelblue",xlab="this is the x-axis",ylab="this is the y-axis",main="Plot of X vs Y") # Add labels
pdf("Figure01.pdf") # Save as pdf, add a path or it will be stored on the project directory
plot(x,y,pch=20, col="forestgreen") # Try different colors?
dev.off() # Close the file using the dev.off function
x=seq(1,10) # Same as x=c(1:10)
x
x=1:10
x
x=seq(-pi,pi,length=50)
y=x

```

## Lab 02

## Indexing Data using \[\]

```{r}
A=matrix(1:16,4,4)
A
A[2,3]
A[c(1,3),c(2,4)]
A[1:3,2:4]
A[1:2,]
A[,1:2]
A[1,]
A[-c(1,3),] # What does -c() do?
A[-c(1,3),-c(1,3,4)]
dim(A) # Dimensions
```

## Loading Data from GitHub (remote)

```{r}
Auto=read.table("https://raw.githubusercontent.com/datageneration/knowledgemining/master/data/Auto.data")
Auto=read.table("https://raw.githubusercontent.com/datageneration/knowledgemining/master/data/Auto.data",header=T,na.strings="?") 
Auto=read.csv("https://raw.githubusercontent.com/datageneration/knowledgemining/master/data/Auto.csv") # read csv file
# Which function reads data faster?

# Try using this simple method
# time1 = proc.time()
# Auto=read.csv("https://raw.githubusercontent.com/datageneration/knowledgemining/master/data/Auto.csv",header=T,na.strings="?")
# proc.time()-time1

# Check on data
dim(Auto)
Auto[1:4,] # select rows
Auto=na.omit(Auto)
dim(Auto) # Notice the difference?
names(Auto)
```

## Load data from ISLR website

```{r Loaddata_from_web}
Auto=read.table("https://www.statlearning.com/s/Auto.data",header=T,na.strings="?")
dim(Auto)
```

## Additional Graphical and Numerical Summaries

```{r}
# plot(cylinders, mpg)
plot(Auto$cylinders, Auto$mpg)
attach(Auto)
plot(cylinders, mpg)
cylinders=as.factor(cylinders)
plot(cylinders, mpg)
plot(cylinders, mpg, col="red")
plot(cylinders, mpg, col="red", varwidth=T)
plot(cylinders, mpg, col="red", varwidth=T,horizontal=T)
plot(cylinders, mpg, col="red", varwidth=T, xlab="cylinders", ylab="MPG")
hist(mpg)
hist(mpg,col=2)
hist(mpg,col=2,breaks=15)
#pairs(Auto)
pairs(~ mpg + displacement + horsepower + weight + acceleration, Auto)
plot(horsepower,mpg)
# identify(horsepower,mpg,name) # Interactive: point and click the dot to identify cases
summary(Auto)
summary(mpg)
```

## Linear Regression

```{r}
ptbu=c("MASS","ISLR")
lapply(ptbu, require, character.only = TRUE)
library(MASS)
library(ISLR)

# Simple Linear Regression

# fix(Boston)
names(Boston)
# lm.fit=lm(medv~lstat)
attach(Boston)
lm.fit=lm(medv~lstat,data=Boston)
attach(Boston)
lm.fit=lm(medv~lstat)
lm.fit
summary(lm.fit)
names(lm.fit)
coef(lm.fit)
confint(lm.fit)
predict(lm.fit,data.frame(lstat=(c(5,10,15))), interval="confidence")
predict(lm.fit,data.frame(lstat=(c(5,10,15))), interval="prediction")
# What is the differnce between "conference" and "prediction" difference?

plot(lstat,medv)
abline(lm.fit)
abline(lm.fit,lwd=3)
abline(lm.fit,lwd=3,col="red")
plot(lstat,medv,col="red")
plot(lstat,medv,pch=16)
plot(lstat,medv,pch="+")
plot(1:20,1:20,pch=1:20)
par(mfrow=c(2,2))
plot(lm.fit)
plot(predict(lm.fit), residuals(lm.fit))
plot(predict(lm.fit), rstudent(lm.fit))
plot(hatvalues(lm.fit))
which.max(hatvalues(lm.fit))
```

## Multiple Linear Regression

```{r}
lm.fit=lm(medv~lstat+age,data=Boston)
summary(lm.fit)
lm.fit=lm(medv~.,data=Boston)
summary(lm.fit)
library(car)
vif(lm.fit)
lm.fit1=lm(medv~.-age,data=Boston)
summary(lm.fit1)
lm.fit1=update(lm.fit, ~.-age)
```

## Non-linear Transformations of the Predictors

```{r}
lm.fit2=lm(medv~lstat+I(lstat^2))
summary(lm.fit2)
lm.fit=lm(medv~lstat)
anova(lm.fit,lm.fit2)
par(mfrow=c(2,2))
plot(lm.fit2)
lm.fit5=lm(medv~poly(lstat,5))
summary(lm.fit5)
summary(lm(medv~log(rm),data=Boston))
```

## Qualitative Predictors

```{r}
# fix(Carseats)
names(Carseats)
lm.fit=lm(Sales~.+Income:Advertising+Price:Age,data=Carseats)
summary(lm.fit)
attach(Carseats)
contrasts(ShelveLoc)
```

## Interaction Terms (including interaction and single effects)

```{r}
summary(lm(medv~lstat*age,data=Boston))
```

# Exploratory Analysis

## Downloading Data-set

```{r}
library(haven)

TEDS_2016 <- haven::read_dta(
  "https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true"
)

names(TEDS_2016)
str(TEDS_2016)
summary(TEDS_2016)
```

## Graphical and Numeral Summaries

```{r}
# 1) Scatter-style plot (only meaningful if x is numeric-ish)
plot(TEDS_2016$Age, TEDS_2016$income)

# 2) Treat Age as categorical (like cylinders)
TEDS_2016$Age_f <- as.factor(TEDS_2016$Age)

plot(TEDS_2016$Age_f, TEDS_2016$income)
plot(TEDS_2016$Age_f, TEDS_2016$income, col="red")
plot(TEDS_2016$Age_f, TEDS_2016$income, col="red", varwidth=TRUE)
plot(TEDS_2016$Age_f, TEDS_2016$income, col="red", varwidth=TRUE, horizontal=TRUE)
plot(TEDS_2016$Age_f, TEDS_2016$income, col="red", varwidth=TRUE,
     xlab="Age group (1–5)", ylab="Income (1–10)")

# 3) Histogram (like hist(mpg))
hist(TEDS_2016$income)
hist(TEDS_2016$income, col=2)
hist(TEDS_2016$income, col=2, breaks=15)

# 4) Pairs plot: pick numeric variables that exist in TEDS
pairs(~ income + age + inequality5 + econworse5 + pubwelf5, data=TEDS_2016)

# 5) Another scatterplot (like plot(horsepower, mpg))
plot(TEDS_2016$age, TEDS_2016$income)

# 6) Summaries
summary(TEDS_2016)
summary(TEDS_2016$income)
```

## Linear Regression

```{r}
names(TEDS_2016)

# Fit the model (no attach needed)
lm.fit <- lm(income ~ age, data = TEDS_2016)

lm.fit
summary(lm.fit)
names(lm.fit)
coef(lm.fit)
confint(lm.fit)

predict(lm.fit, data.frame(age = c(25, 40, 60)), interval = "confidence")
predict(lm.fit, data.frame(age = c(25, 40, 60)), interval = "prediction")

plot(TEDS_2016$age, TEDS_2016$income)
abline(lm.fit)

abline(lm.fit, lwd=3)
abline(lm.fit, lwd=3, col="red")

plot(TEDS_2016$age, TEDS_2016$income, col="red")
plot(TEDS_2016$age, TEDS_2016$income, pch=16)
plot(TEDS_2016$age, TEDS_2016$income, pch="+")

par(mfrow=c(2,2))
plot(lm.fit)

plot(predict(lm.fit), residuals(lm.fit))
plot(predict(lm.fit), rstudent(lm.fit))
plot(hatvalues(lm.fit))
which.max(hatvalues(lm.fit))

lm.edu <- lm(income ~ as.factor(Edu), data = TEDS_2016)
summary(lm.edu)

plot(as.factor(TEDS_2016$Edu), TEDS_2016$income, col="red", varwidth=TRUE)
```

## Multiple Linear Regression

```{r}
install.packages("car")
library(car)

# (Assumes TEDS_2016 is already loaded)
# Optional: remove rows with missing values for variables used
TEDS_lm <- subset(TEDS_2016,
                  !is.na(income) & !is.na(age) & !is.na(edu) &
                  !is.na(female) & !is.na(whitecollar))

# 1) Multiple regression like: medv ~ lstat + age
lm.fit <- lm(income ~ age + edu, data = TEDS_lm)
summary(lm.fit)

# 2) "All predictors" version (safe approach: choose a set of numeric predictors)
lm.fit_all <- lm(income ~ age + edu + female + whitecollar + inequality5 + econworse5 + pubwelf5,
                 data = TEDS_lm)
summary(lm.fit_all)

# VIF for multicollinearity (works on lm objects)
vif(lm.fit_all)

# 3) Remove one variable (like ~ . - age)
lm.fit1 <- update(lm.fit_all, ~ . - age)
summary(lm.fit1)

# same thing, just showing the update call directly (like your last line)
lm.fit1b <- update(lm.fit_all, ~ . - age)
```

## Non-linear Transformations of the Predictors

```{r}
# Assumes TEDS_2016 is already loaded

# Keep only complete cases for variables we use
TEDS_poly <- subset(TEDS_2016, !is.na(income) & !is.na(age) & age > 0)

# Quadratic model: income ~ age + age^2
lm.fit2 <- lm(income ~ age + I(age^2), data = TEDS_poly)
summary(lm.fit2)

# Linear model: income ~ age
lm.fit <- lm(income ~ age, data = TEDS_poly)

# Compare linear vs quadratic (nested models)
anova(lm.fit, lm.fit2)

# Diagnostic plots for quadratic model
par(mfrow = c(2,2))
plot(lm.fit2)

# 5th-degree polynomial model
lm.fit5 <- lm(income ~ poly(age, 5), data = TEDS_poly)
summary(lm.fit5)

# Log-transform example (like log(rm) in Boston)
# Here: income ~ log(age)
summary(lm(income ~ log(age), data = TEDS_poly))
```

## Qualitative Predictors

```{r}
# Make sure dataset is loaded
# library(haven)
# TEDS_2016 <- read_dta("...")

# Remove missing values for variables used
TEDS_int <- subset(TEDS_2016,
                   !is.na(income) & !is.na(age) &
                   !is.na(edu) & !is.na(inequality5) &
                   !is.na(econworse5))

# Convert Edu to factor (like ShelveLoc)
TEDS_int$Edu_f <- as.factor(TEDS_int$Edu)

# Full model with interactions
lm.fit <- lm(income ~ age + edu + inequality5 + econworse5 +
               age:inequality5 + edu:econworse5,
             data = TEDS_int)

summary(lm.fit)
```

## Interactive Terms

```{r}
# Remove missing values for variables used
TEDS_int <- subset(TEDS_2016,
                   !is.na(income) &
                   !is.na(age) &
                   !is.na(inequality5))

summary(lm(income ~ age * inequality5, data = TEDS_int))


```
