---
title: "Assignment 3 - March 24th, 2026"
subtitle: EPPS 6323 Knowledge Mining
format: 
  html:
    code-fold: true
    code-summary: "Show the Code"
---

## Lab Prep

Data Base

```{r}
library(haven)

TEDS_2016 <- read_stata(
  "https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_20016.dta?raw=true"
  )

```

ML01.R

```{r}
# ============================================================================
# Workshop: Machine Learning with Survey Data (TEDS 2016)
# EPPS 6323 Knowledge Mining
# Karl Ho, University of Texas at Dallas
# ============================================================================

# ----------------------------------------------------------------------------
# 0. Setup: Packages and Data
# ----------------------------------------------------------------------------

library(haven)        # Read Stata files
library(tidyverse)    # Data wrangling and visualization
library(GGally)       # Pairs plots
library(cluster)      # Clustering
library(factoextra)   # Visualize clusters and PCA
library(rpart)        # Decision trees
library(rpart.plot)   # Plot decision trees
library(randomForest) # Random forest
library(caret)        # Model training and evaluation
library(e1071)        # SVM and Naive Bayes

# Load data
TEDS_2016 <- read_stata("https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true")


# ============================================================================
# PART I: EXPLORATORY DATA ANALYSIS
# ============================================================================

# ----------------------------------------------------------------------------
# Q1. Data overview
# ----------------------------------------------------------------------------

dim(TEDS_2016)
names(TEDS_2016)
summary(TEDS_2016)

# ----------------------------------------------------------------------------
# Q2. Missing data assessment
# ----------------------------------------------------------------------------

TEDS_2016 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  arrange(desc(n_missing))

# Visualize missingness
TEDS_2016 %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), 
               names_to = "variable", 
               values_to = "n_missing") %>%
  filter(n_missing > 0) %>%
  ggplot(aes(x = reorder(variable, n_missing), y = n_missing)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Missing Values by Variable",
       x = NULL, y = "Count of Missing Values") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q3. Variable recoding
# ----------------------------------------------------------------------------

teds <- TEDS_2016 %>%
  mutate(
    vote = factor(votetsai, 
                  levels = c(0, 1), 
                  labels = c("Other", "Tsai")),
    gender = factor(female, 
                    levels = c(0, 1), 
                    labels = c("Male", "Female")),
    Tondu = as.factor(Tondu),
    Party = as.factor(Party)
  ) %>%
  dplyr::select(vote, gender, age, edu, income, 
         Taiwanese, Econ_worse, Tondu, Party, DPP) %>%
  drop_na()

glimpse(teds)

# ----------------------------------------------------------------------------
# Q4. Vote choice distribution
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = vote, fill = vote)) +
  geom_bar() +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice Distribution",
       subtitle = "TEDS 2016: Tsai Ing-wen vs. Other Candidates",
       x = "Vote Choice", y = "Count") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")

# ----------------------------------------------------------------------------
# Q5. Age distribution
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(title = "Age Distribution of Respondents",
       x = "Age", y = "Count") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q6. Vote by gender
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = gender, fill = vote)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice by Gender",
       x = "Gender", y = "Proportion",
       fill = "Vote") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q7. Age by vote choice
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = vote, y = age, fill = vote)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Age Distribution by Vote Choice",
       x = "Vote Choice", y = "Age") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")

# ----------------------------------------------------------------------------
# Q8. Correlation matrix
# ----------------------------------------------------------------------------

teds %>%
  dplyr::select(age, edu, income, Taiwanese, Econ_worse, DPP) %>%
  cor(use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "cor") %>%
  ggplot(aes(x = var1, y = var2, fill = cor)) +
  geom_tile() +
  geom_text(aes(label = round(cor, 2)), size = 6) +
  scale_fill_gradient2(low = "#e74c3c", mid = "white", high = "#2980b9",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix of Numeric Variables",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 18) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ----------------------------------------------------------------------------
# Q9. Pairs plot
# ----------------------------------------------------------------------------

teds %>%
  dplyr::select(age, edu, income, Taiwanese, Econ_worse, vote) %>%
  ggpairs(aes(color = vote, alpha = 0.5),
          upper = list(continuous = wrap("cor", size = 4)),
          lower = list(continuous = wrap("points", size = 0.5))) +
  scale_color_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  theme_minimal(base_size = 14)


# ============================================================================
# PART II: UNSUPERVISED LEARNING
# ============================================================================

# ----------------------------------------------------------------------------
# Prepare data for clustering
# ----------------------------------------------------------------------------

teds_numeric <- teds %>%
  dplyr::select(age, edu, income, Taiwanese, Econ_worse, DPP) %>%
  scale() %>%
  as.data.frame()

head(teds_numeric)

# ----------------------------------------------------------------------------
# Q10. Choosing k: Elbow method
# ----------------------------------------------------------------------------

fviz_nbclust(teds_numeric, kmeans, method = "wss", k.max = 10) +
  labs(title = "Elbow Method: Optimal Number of Clusters") +
  theme_minimal(base_size = 18)

# Silhouette method
fviz_nbclust(teds_numeric, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette Method: Optimal Number of Clusters") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q11. K-Means clustering
# ----------------------------------------------------------------------------

set.seed(6323)
km_result <- kmeans(teds_numeric, centers = 3, nstart = 25)

table(km_result$cluster)
round(km_result$centers, 2)

# Visualize clusters
fviz_cluster(km_result, data = teds_numeric,
             geom = "point", 
             ellipse.type = "convex",
             palette = c("#e74c3c", "#2980b9", "#27ae60"),
             ggtheme = theme_minimal(base_size = 18)) +
  labs(title = "K-Means Clustering of TEDS 2016 Voters (k = 3)")

# ----------------------------------------------------------------------------
# Q12. Cluster profiling
# ----------------------------------------------------------------------------

teds$cluster <- as.factor(km_result$cluster)

teds %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    mean_age = round(mean(age), 1),
    mean_edu = round(mean(edu), 1),
    mean_income = round(mean(income), 1),
    pct_Taiwanese = round(mean(Taiwanese) * 100, 1),
    pct_Econ_worse = round(mean(Econ_worse) * 100, 1),
    mean_DPP = round(mean(DPP), 2),
    pct_Tsai = round(mean(vote == "Tsai") * 100, 1)
  )

# ----------------------------------------------------------------------------
# Q13. Clusters vs. vote choice
# ----------------------------------------------------------------------------

ggplot(teds, aes(x = cluster, fill = vote)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("Other" = "#2980b9", "Tsai" = "#27ae60")) +
  labs(title = "Vote Choice Distribution by Cluster",
       x = "Cluster", y = "Proportion",
       fill = "Vote") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q14. Principal Component Analysis
# ----------------------------------------------------------------------------

pca_result <- prcomp(teds_numeric, scale. = TRUE)
summary(pca_result)

# Scree plot
fviz_eig(pca_result, addlabels = TRUE) +
  labs(title = "Scree Plot: Variance Explained by Each PC") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q15. PCA variable contributions
# ----------------------------------------------------------------------------

fviz_pca_var(pca_result, 
             col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE) +
  labs(title = "PCA: Variable Contributions") +
  theme_minimal(base_size = 18)

# PCA individuals by vote
fviz_pca_ind(pca_result,
             geom = "point",
             col.ind = teds$vote,
             palette = c("#2980b9", "#27ae60"),
             addEllipses = TRUE,
             legend.title = "Vote") +
  labs(title = "PCA: Voters Projected onto First Two PCs") +
  theme_minimal(base_size = 18)


# ============================================================================
# PART III: SUPERVISED LEARNING
# ============================================================================

# ----------------------------------------------------------------------------
# Q16. Train-test split
# ----------------------------------------------------------------------------

set.seed(6323)

train_index <- createDataPartition(teds$vote, p = 0.7, list = FALSE)
train_data <- teds[train_index, ] %>% dplyr::select(-cluster)
test_data  <- teds[-train_index, ] %>% dplyr::select(-cluster)

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
prop.table(table(train_data$vote))

# ----------------------------------------------------------------------------
# Q17. Logistic regression
# ----------------------------------------------------------------------------

logit_model <- glm(vote ~ age + gender + edu + income + 
                     Taiwanese + Econ_worse + DPP,
                   data = train_data, 
                   family = binomial)

summary(logit_model)

# Evaluation
logit_probs <- predict(logit_model, newdata = test_data, type = "response")
logit_pred <- ifelse(logit_probs > 0.5, "Tsai", "Other")
logit_pred <- factor(logit_pred, levels = c("Other", "Tsai"))

confusionMatrix(logit_pred, test_data$vote)

# ----------------------------------------------------------------------------
# Q18. Decision tree
# ----------------------------------------------------------------------------

tree_model <- rpart(vote ~ age + gender + edu + income + 
                      Taiwanese + Econ_worse + DPP,
                    data = train_data,
                    method = "class")

rpart.plot(tree_model, 
           type = 4, 
           extra = 104,
           main = "Decision Tree: Predicting Vote Choice",
           box.palette = "BuGn",
           cex = 1.2)

# Evaluation
tree_pred <- predict(tree_model, newdata = test_data, type = "class")
confusionMatrix(tree_pred, test_data$vote)

# ----------------------------------------------------------------------------
# Q19. Random forest
# ----------------------------------------------------------------------------

set.seed(6323)

rf_model <- randomForest(vote ~ age + gender + edu + income + 
                           Taiwanese + Econ_worse + DPP,
                         data = train_data,
                         ntree = 500,
                         importance = TRUE)

print(rf_model)

# Evaluation
rf_pred <- predict(rf_model, newdata = test_data)
confusionMatrix(rf_pred, test_data$vote)

# ----------------------------------------------------------------------------
# Q20. Variable importance
# ----------------------------------------------------------------------------

importance_df <- as.data.frame(importance(rf_model)) %>%
  rownames_to_column("variable") %>%
  arrange(desc(MeanDecreaseGini))

ggplot(importance_df, aes(x = reorder(variable, MeanDecreaseGini), 
                           y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Random Forest: Variable Importance",
       subtitle = "Mean Decrease in Gini Index",
       x = NULL, y = "Importance") +
  theme_minimal(base_size = 18)

# ----------------------------------------------------------------------------
# Q21. Model comparison
# ----------------------------------------------------------------------------

results <- tibble(
  Model = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(
    confusionMatrix(logit_pred, test_data$vote)$overall["Accuracy"],
    confusionMatrix(tree_pred, test_data$vote)$overall["Accuracy"],
    confusionMatrix(rf_pred, test_data$vote)$overall["Accuracy"]
  )
)

results %>% arrange(desc(Accuracy))

# Comparison plot
ggplot(results, aes(x = reorder(Model, Accuracy), y = Accuracy, fill = Model)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(Accuracy * 100, 1), "%")), 
            hjust = -0.1, size = 7) +
  coord_flip() +
  scale_fill_manual(values = c("#e74c3c", "#2980b9", "#27ae60")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(title = "Model Comparison: Prediction Accuracy",
       x = NULL, y = "Accuracy") +
  theme_minimal(base_size = 18) +
  theme(legend.position = "none")

```

## Multiple Regression Model

Data Check

```{r}
library(haven)
library(tidyverse)
library(car)

TEDS_2016 <- read_stata(
  "https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true"
)

TEDS_2016_clean <- TEDS_2016 %>%
  mutate(across(where(is.labelled), as_factor))

names(TEDS_2016_clean)
```

Model

```{r}
library(haven)
library(tidyverse)
library(car)

TEDS_2016 <- read_stata(
  "https://github.com/datageneration/home/blob/master/DataProgramming/data/TEDS_2016.dta?raw=true"
)

# Keep only the variables of interest and remove missing values
TEDS_reg <- TEDS_2016 %>%
  select(Age, Edu, income) %>%
  drop_na()

# Fit model with two predictors
lm.fit <- lm(Age ~ Edu + income, data = TEDS_reg)
summary(lm.fit)

# VIF for multicollinearity
vif(lm.fit)

# Model removing one predictor
lm.fit1 <- update(lm.fit, ~ . - Age)
summary(lm.fit1)

# Same thing using update()
lm.fit1 <- update(lm.fit, ~ . - Edu)
summary(lm.fit1)

# And another update()
lm.fit1 <- update(lm.fit, ~ . - income)
summary(lm.fit1)
```

Plot

```{r}
## Creating a function: regplot 
## Combine the lm, plot and abline functions to create a regression fit plot 
regplot <- function(x, y) {
  fit <- lm(y ~ x)
  plot(x, y,
       pch = 19,
       col = "darkgray",
       xlab = deparse(substitute(x)),
       ylab = deparse(substitute(y)),
       main = "Regression Plot")
  abline(fit, col = "red", lwd = 2)
}

regplot(TEDS_reg$Age, TEDS_reg$Edu)
regplot(TEDS_reg$Age, TEDS_reg$income)
regplot(TEDS_reg$Edu, TEDS_reg$income)
```

## Reflection

The lines look neat enough, but something feels off. The relationships are weak, drifting, like they’re trying to say something but never quite get there. The dependent variable (political leaning) isn’t behaving like a clean, continuous number. It’s more like a set of categories, several distinct buckets rather than a smooth scale, so forcing a straight-line regression onto it is a bit like using a ruler to measure a staircase. You can do it, but it won’t capture what’s actually going on.

The model struggles not because the code is wrong, but because the setup doesn’t quite match the data. A better move would be switching gears, something that can fit the situation more naturally. Also, tossing in more meaningful predictors (things tied more directly to political attitudes) would probably help. Age, education, income are fine, but on their own they’re kind of blunt instruments. We need sharper tools if you want the predictions to land.
