---
title: "Base Perspective Diagrams"
format: html
---

**Base 2D Plot:**

This plot is made using ggplot2 and viridis to give the plot a clean look. No real data is being used in the plot.

```{r}
# Install packages if needed
# Load libraries
library(ggplot2)
library(viridis)

# Example data
set.seed(123)
data <- data.frame(
  x = rnorm(200),
  y = rnorm(200),
  z = rnorm(200)  # this will control color/size for perspective effect
)

# Perspective-like scatter plot
ggplot(data, aes(x = x, y = y, color = z, size = z)) +
  geom_point(alpha = 0.8) +                      # points with transparency
  scale_color_viridis(option = "D") +            # viridis color scale
  labs(
    title = "2D Perspective Scatter Plot",
    x = "X Axis",
    y = "Y Axis",
    color = "Z Value",
    size = "Z Value"
  ) +
  theme_minimal()


```

**Plot with Work Color Palette:**

For work, I incorporate a custom palette for my plots.

```{r}
library(ggplot2)

# Example data
set.seed(123)
data <- data.frame(
  x = rnorm(200),
  y = rnorm(200),
  z = rnorm(200)  # this will control color/size for perspective effect
)

# Your custom palette
work_viridis <- colorRampPalette(c("#021C49","#1F3657","#3C5C7C",
                                   "#7FA0C0", "#E9EEF3", "#F3F7FF"))
custom_colors <- work_viridis(256)  # generate 256 colors

# Perspective-like scatter plot with custom palette
ggplot(data, aes(x = x, y = y, color = z, size = z)) +
  geom_point(alpha = 0.8) +                        # points with transparency
  scale_color_gradientn(colors = custom_colors) +  # use custom palette
  labs(
    title = "2D Perspective Scatter Plot with Custom Palette",
    x = "X Axis",
    y = "Y Axis",
    color = "Z Value",
    size = "Z Value"
  ) +
  theme_minimal()

```

**Plot with Primary & Secondary Work Palettes to Create Contrasts:**

```{r}
library(ggplot2)

# Example data
set.seed(123)
data <- data.frame(
  x = rnorm(200),
  y = rnorm(200),
  z = rnorm(200)  # this will control color/size for perspective effect
)

# Theme palette
work_viridis <- colorRampPalette(c("#021C49","#1F3657","#3C5C7C",
                                   "#7FA0C0", "#E9EEF3", "#F3F7FF"))
theme_colors <- work_viridis(256)
bg_light <- theme_colors[256]   # light background
line_dark <- theme_colors[1]  # dark text/lines

# Data palette
work2_viridis <- colorRampPalette(c("#55401C", "#7f602A", "#A98038","#D4B57F","#F1E6Df"))
data_colors <- work2_viridis(256)

# 2D Perspective-like scatter plot
ggplot(data, aes(x = x, y = y, color = z, size = z)) +
  geom_point(alpha = 0.8) +
  scale_color_gradientn(colors = data_colors) +  # data points use work2_viridis
  labs(
    title = "2D Perspective Scatter Plot with Dual Palettes",
    x = "X Axis",
    y = "Y Axis",
    color = "Z Value",
    size = "Z Value"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = bg_light, color = NA),
    panel.background = element_rect(fill = bg_light, color = NA),
    panel.grid.major = element_line(color = line_dark, linewidth = 0.3),
    panel.grid.minor = element_line(color = line_dark, linewidth = 0.1),
    axis.title = element_text(color = line_dark, size = 12),
    axis.text = element_text(color = line_dark, size = 10),
    plot.title = element_text(color = line_dark, size = 14, face = "bold"),
    legend.title = element_text(color = line_dark),
    legend.text  = element_text(color = line_dark)
  )

```

**Base 3D Plot:**

This plot is made using ggplot2 and viridis to give the plot a clean look. No real data is being used in the plot.

```{r}
# Install packages if needed
# Load libraries
library(plotly)
library(viridis)

# Example data
set.seed(123)
data <- data.frame(
  x = rnorm(200),
  y = rnorm(200),
  z = rnorm(200)
)

# Generate a viridis color palette based on z
colors <- viridis(200)

# 3D Scatter Plot
plot_ly(
  data, 
  x = ~x, y = ~y, z = ~z, 
  type = "scatter3d", 
  mode = "markers",
  marker = list(
    size = 5,
    color = ~z,          # color by z
    colorscale = 'Viridis',  # viridis colorscale
    showscale = TRUE
  )
) %>%
  layout(
    scene = list(
      xaxis = list(title = "X Axis"),
      yaxis = list(title = "Y Axis"),
      zaxis = list(title = "Z Axis")
    ),
    title = "3D Scatter Plot with Viridis Colors"
  )

```

**Plot with Work Color Palette:**

For work, I incorporate a custom palette for my plots.

```{r}
library(plotly)

# Example data
set.seed(123)
data <- data.frame(
  x = rnorm(200),
  y = rnorm(200),
  z = rnorm(200)
)

# Custom theme palette
work_viridis <- colorRampPalette(c("#021C49","#1F3657","#3C5C7C",
                                   "#7FA0C0", "#E9EEF3", "#F3F7FF"))
custom_colors <- work_viridis(256)  # generate 256 colors

# Convert palette to Plotly colorscale format
plotly_colorscale <- lapply(seq_along(custom_colors), function(i) {
  c((i-1)/(length(custom_colors)-1), custom_colors[i])
})

# 3D Scatter Plot with custom colorscale
plot_ly(
  data, 
  x = ~x, y = ~y, z = ~z, 
  type = "scatter3d", 
  mode = "markers",
  marker = list(
    size = 5,
    color = ~z,                  # color by z
    colorscale = plotly_colorscale,  # use custom palette
    showscale = TRUE
  )
) %>%
  layout(
    scene = list(
      xaxis = list(title = "X Axis"),
      yaxis = list(title = "Y Axis"),
      zaxis = list(title = "Z Axis")
    ),
    title = "3D Scatter Plot with Custom Theme Palette"
  )

```

**Plot with Primary & Secondary Work Palettes to Create Contrasts:**

```{r}
library(plotly)

# Example data
set.seed(123)
data <- data.frame(
  x = rnorm(200),
  y = rnorm(200),
  z = rnorm(200)
)

# Theme palette (for background, grids, axes)
work_viridis <- colorRampPalette(c("#021C49","#1F3657","#3C5C7C",
                                   "#7FA0C0", "#E9EEF3", "#F3F7FF"))
theme_colors <- work_viridis(256)
bg_light <- theme_colors[256]   # light background
line_dark <- theme_colors[1]  # dark axes/grid lines

# Data palette (for points)
work2_viridis <- colorRampPalette(c("#55401C", "#7f602A", "#A98038","#D4B57F","#F1E6Df"))
data_colors <- work2_viridis(256)

# Convert data palette to Plotly colorscale format
data_colorscale <- lapply(seq_along(data_colors), function(i) {
  c((i-1)/(length(data_colors)-1), data_colors[i])
})

# 3D Scatter Plot with dual palettes
plot_ly(
  data, 
  x = ~x, y = ~y, z = ~z, 
  type = "scatter3d", 
  mode = "markers",
  marker = list(
    size = 5,
    color = ~z,                   # color by z
    colorscale = data_colorscale, # use work2_viridis for points
    showscale = TRUE
  )
) %>%
  layout(
    scene = list(
      xaxis = list(title = "X Axis", backgroundcolor = bg_light, gridcolor = line_dark, linecolor = line_dark, showbackground = TRUE),
      yaxis = list(title = "Y Axis", backgroundcolor = bg_light, gridcolor = line_dark, linecolor = line_dark, showbackground = TRUE),
      zaxis = list(title = "Z Axis", backgroundcolor = bg_light, gridcolor = line_dark, linecolor = line_dark, showbackground = TRUE)
    ),
    paper_bgcolor = bg_light,  # overall plot background
    title = "3D Scatter Plot with Dual Palettes"
  )
```
